---
title: "Bayesian estimation for NHPP using `rstan`"
author: "Miao Cai <miao.cai@slu.edu>"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2: 
    number_sections: true
    toc: true
header-includes:
  - \usepackage{soul}
  - \usepackage{float}
  - \usepackage{xcolor}
  - \usepackage{setspace}\linespread{1.4}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

\clearpage

The `Stan` code I used previously cannot estimate trips without any critical event. In this document, I will try to solve this issue.

```{r}
source("functions/NHPP_functions.R")
require(rstan)
rstan_options(auto_write = TRUE)
```

\clearpage

# Reproduce the error

```{r}
plptau = '
functions{
  real nhpp_log(vector t, real beta, real theta, real tau){
    vector[num_elements(t)] loglik_part;
    real loglikelihood;
    for (i in 1:num_elements(t)){
      loglik_part[i] = log(beta) - beta*log(theta) + (beta - 1)*log(t[i]);
    }
    loglikelihood = sum(loglik_part) - (tau/theta)^beta;
    return loglikelihood;
  }
}
data {
  int<lower=0> N; //total # of obs
  int<lower=0> K; //total # of groups
  vector<lower=0>[K] tau;//truncated time
  vector<lower=0>[N] event_time; //failure time
  int s[K]; //group sizes
}
parameters{
  real<lower=0> beta;
  real<lower=0> theta;
}
model{
  int position;
  position = 1;
  for (k in 1:K){
    segment(event_time, position, s[k]) ~ nhpp(beta, theta, tau[k]);
    position = position + s[k];
  }
//PRIORS
  beta ~ gamma(1, 1);
  theta ~ gamma(1, 0.01);
}
'
```

\clearpage
## An example that works

Firstly, I create a example with sufficiently long time intervals so that there are at least 1 event in each time interval.

```{r}
df0 = list(
  N = 38L, K = 10L, 
  tau = c(21.269, 18.109, 19.468, 19.89, 18.247, 19.048, 19.957, 21.006, 
17.524, 19.475), 
  event_time = c(5.045, 14.921, 18.566, 7.265, 10.51, 12.155, 16.262, 17.738, 
17.763, 16.059, 18.371, 10.393, 11.787, 5.088, 10.144, 11.646, 
13.274, 15.233, 16.345, 17.583, 15.266, 15.391, 16.355, 17.79, 
7.729, 13.906, 14.287, 12.012, 18.662, 5.654, 5.727, 8.144, 11.608, 
14.756, 14.933, 16.088, 16.45, 18.876), 
  s = c(3L, 6L, 2L, 2L, 7L, 4L, 3L, 2L, 6L, 3L)
)
df0
```

`Stan` works out well and produce estimates close to the true parameters $\beta = 2, \theta = 10$.


```{r}
fitplp <- stan(
  model_code=plptau, model_name="NHPP", data=df0, 
  iter=1000, warmup = 500, chains=1, seed = 123, refresh = 0
)
fitplp
```

\clearpage

## An example that does not work

However, if there is no event in any of the time intervals, the `Stan` code will not work. For example:

```{r}
df1 = list(
  N = 8L, K = 10L, 
  tau = c(9.9785, 7.3146, 10.0518, 10.2853, 10.2621, 8.4175, 10.7142, 12.0679, 10.6844, 8.2966), 
  event_time = c(6.7346, 8.1608, 4.4621, 6.5118, 7.9538, 11.2332, 11.6206, 11.9121), 
  s = c(0L, 0L, 2L, 1L, 2L, 0L, 0L, 3L, 0L, 0L)
)
df1
```

```{r error=TRUE}
fitplp <- stan(
  model_code=plptau, model_name="NHPP", data=df1, 
  iter=1000, warmup = 500, chains=1, seed = 123, refresh = 0
)
```

\clearpage

# Update code

Here is a solution by just add one line of code in the `for` loop in `model` chunk: `if(s[k] == 0) continue;`, which is provided by [jjramsey](https://discourse.mc-stan.org/u/jjramsey) on [Stan discourse](https://discourse.mc-stan.org/t/zero-length-vector-for-ragged-data-structures/9605/2).

\linespread{1}
```{r}
plptau1 = '
functions{
  real nhpp_log(vector t, real beta, real theta, real tau){
    vector[num_elements(t)] loglik_part;
    real loglikelihood;
    for (i in 1:num_elements(t)){
      loglik_part[i] = log(beta) - beta*log(theta) + (beta - 1)*log(t[i]);
    }
    loglikelihood = sum(loglik_part) - (tau/theta)^beta;
    return loglikelihood;
  }
}
data {
  int<lower=0> N; //total # of obs
  int<lower=0> K; //total # of groups
  vector<lower=0>[K] tau;//truncated time
  vector<lower=0>[N] event_time; //failure time
  int s[K]; //group sizes
}
parameters{
  real<lower=0> beta;
  real<lower=0> theta;
}
model{
  int position;
  position = 1;
  for (k in 1:K){
    if(s[k] == 0) continue;
    segment(event_time, position, s[k]) ~ nhpp(beta, theta, tau[k]);
    position = position + s[k];
  }
//PRIORS
  beta ~ gamma(1, 1);
  theta ~ gamma(1, 0.01);
}
'
```
\linespread{1.4}
\clearpage

## A small example

```{r}
df1 = list(
  N = 8L, K = 10L, 
  tau = c(9.9785, 7.3146, 10.0518, 10.2853, 10.2621, 8.4175, 10.7142, 12.0679, 10.6844, 8.2966), 
  event_time = c(6.7346, 8.1608, 4.4621, 6.5118, 7.9538, 11.2332, 11.6206, 11.9121), 
  s = c(0L, 0L, 2L, 1L, 2L, 0L, 0L, 3L, 0L, 0L)
)
```

```{r error=TRUE}
fitplp <- stan(
  model_code=plptau1, model_name="NHPP", data=df1, 
  iter=1000, warmup = 500, chains=1, seed = 123, refresh = 0
)
fitplp
```

## Larger sample size

```{r}
set.seed(123)
nhpp_5 = sim_mul_plp_tau(n_shift = 1000, shift_len_mean = 10, shift_len_sd = 2)
datstan1 = list(
    N = nrow(nhpp_5$event_dat),
    K = nrow(nhpp_5$start_end_dat),
    tau = nhpp_5$start_end_dat$end_time,
    event_time = nhpp_5$event_dat$event_time,
    s = nhpp_5$shift_length
)

fitplp <- stan(
  model_code=plptau1, model_name="NHPP", data=datstan1, 
  iter=2000, warmup = 1000, chains=1, seed = 123, refresh = 0
)
fitplp
```


