---
title: "Bayesian hierarchical models for NHPP using `rstan`"
author: "Miao Cai <miao.cai@slu.edu>"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2: 
    number_sections: true
    toc: true
header-includes:
  - \usepackage{soul}
  - \usepackage{float}
#  - \usepackage[table]{xcolor}
  - \usepackage{setspace}\doublespacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Model setting

Let $T_{d, s, i}$ denote the time to the $d$-th driver's $s$-th shift's $i$-th critical event. The total number critical events of $d$-th driver's $s$-th shift is $n_{d,s}$. The ranges of these notations are:

- $i = 1, 2, \cdots, n_{d, S_d}$,
- $s = 1, 2, \cdots, S_d$,
- $d = 1, 2, \cdots, D$.

We assume the times of critical events within the $d$-th driver's  $s$-th shift were generated from a non-homogeneous Poisson process (NHPP) with a power law process (PLP), with a fix rate parameter $\beta$ and varying scale parameters $\theta_{d, s}$ across drivers. The data generating process is then:

$$
\begin{aligned}
T_{d, s, 1}, T_{d, s, 2}, \cdots , T_{d, s, n_{d, s}} & \sim \text{PLP}(\beta, \theta_{d, s})\\
\beta & \sim \text{Gamma}(1, 1)\\
\log\theta_{d, s} &= \gamma_{0d} + \gamma_{1}x_{d, s, 1} + \gamma_{2}x_{d, s, 2} + \cdots + \gamma_{k}x_{d, s, k}\\
\gamma_{01}, \gamma_{02}, \cdots, \gamma_{0D} & \sim \text{i.i.d. }N(\mu_0, \sigma_0^2)\\
\gamma_1, \gamma_2, \cdots, \gamma_k & \sim \text{i.i.d. }N(0, 10^2)\\
\mu_0 &\sim N(0, 10^2) \\
\sigma_0 &\sim \text{Gamma}(1, 1)
\end{aligned}
$$

\clearpage
# Simulating data

\linespread{1.5}

1. Random intercepts $\gamma_{01}, \gamma_{02}, \cdots, \gamma_{0D}$. The standard deviation of $\mu_0$ was intentionally set to small number 2 to make $\theta_{d, s}$ fall into a reasonably small range. If I otherwise set it as 10,  $\theta_{d, s}$ may be more than $10^5$ due to the exponentiation, which may not be realistic in real-life data.
$$
\begin{aligned}
\mu_0 = 1 &, \quad \sigma_0 = 1 \\
\sigma_0 &\sim \text{Gamma}(1, 1)\\
\gamma_{01}, \gamma_{02}, \cdots, \gamma_{0D} & \sim \text{i.i.d. }N(\mu_0, \sigma_0^2)
\end{aligned}
$$

2. fixed parameters: 3 fixed parameters $\gamma_1, \gamma_2, \gamma_3$. $$\gamma_1, \gamma_2, \gamma_3 \sim \text{i.i.d. }N(0, 10^2)$$

3. The number of observations in the $d$-th driver: $N_{d}$. $$N_{d} \sim \text{Poisson}(100)$$

4. Data: 3 predictor variables $x_{d, s, 1}, x_{d, s, 2}, x_{d, s, 3}$.
$$
\begin{aligned}
x_{d, s, 1} &\sim \text{N}(0, 10)\\
x_{d, s, 2} &\sim \text{Gamma}(10, 2)\\
x_{d, s, 3} &\sim \text{Poisson}(3.5)
\end{aligned}
$$

5. Scale parameters of a NHPP (random effects): $\theta_{d, s}$.
$$
\theta_{d, s} = \text{EXP}(\gamma_{0d} + \gamma_{1}x_{d, s, 1} + \gamma_{2}x_{d, s, 2} + \gamma_{k}x_{d, s, 3})
$$

6. Shape parameter of a NHPP (fixed effect): $\beta  \sim \text{Gamma}(1, 1)$. Set $$\beta = 1.5$$

7. Simulate a NHPP based on $\beta$ and $\theta_{d, s}$. $$T_{d, s, 1}, T_{d, s, 2}, \cdots , T_{d, s, n_{d, s}} \sim \text{PLP}(\beta, \theta_{d, s})$$

\clearpage
\singlespacing
```{r parameters}
pacman::p_load(rstan, tidyverse, data.table)
source("functions/NHPP_functions.R")

#set.seed(123)
D = 10 # the number of drivers
K = 3 # the number of predictor variables

# 1. Random-effect intercepts
# hyperparameters
mu0 = 1
sigma0 = 1
r_0D = rnorm(D, mean = mu0, sd = sigma0)

# 2. Fixed-effects parameters
R_K = rnorm(K, mean = 0, sd = 1)

# 3. The number of observations in the $d$-th driver: $N_{d}$
N_K = rpois(D, 10)

# 4. Generate data: x_1, x_2, .. x_K
sim1 = function(n = 10){
  x1 = rnorm(n, 0, 5)
  x2 = rgamma(n, 5, 1)
  x3 = rpois(n, 3.5)
  return(data.frame(x1, x2, x3))
}
simXD = function(ndrivers = D){
  XD = rep(list(data.frame()), ndrivers)
  for (i in 1:D) {
    XD[[i]] = sim1(N_K[i])
  }
  return(data.table::rbindlist(XD, idcol = "driver"))
}
X = simXD()

# 5. Scale parameters of a NHPP
# 5a. parameter matrix: P
N_D = X[,.N,driver][["N"]]# N by driver
N_all = sum(N_D) # total N
P = cbind(r0 = rep(r_0D, N_D), 
          t(replicate(N_all, R_K)))
M_logtheta = P*X

theta = exp(rowSums(M_logtheta))
beta = 1.5

```
\doublespacing


```{r NHPPdata}

```

\clearpage
# Estimation


\linespread{1.4}
```{r}
plptauML = '
functions{
  real nhpp_log(vector t, real beta, real theta, real tau){
    vector[num_elements(t)] loglik_part;
    real loglikelihood;
    for (i in 1:num_elements(t)){
      loglik_part[i] = log(beta) - beta*log(theta) + (beta - 1)*log(t[i]);
    }
    loglikelihood = sum(loglik_part) - (tau/theta)^beta;
    return loglikelihood;
  }
}
data {
  int<lower=0> N; //total # of obs
  int<lower=0> K; //total # of groups
  vector<lower=0>[K] tau;//truncated time
  vector<lower=0>[N] event_time; //failure time
  int s[K]; //group sizes
}
parameters{
  real<lower=0> beta;
  real<lower=0> theta;
}
model{
  int position;
  position = 1;
  for (k in 1:K){
    segment(event_time, position, s[k]) ~ nhpp(beta, theta, tau[k]);
    position = position + s[k];
  }
//PRIORS
  beta ~ gamma(1, 1);
  theta ~ gamma(1, 0.01);
}
'
```
\doublespacing

