---
title: "Bayesian hierarchical models for NHPP using `rstan`"
author: "Miao Cai <miao.cai@slu.edu>"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2: 
    number_sections: true
    toc: true
header-includes:
  - \usepackage{soul}
  - \usepackage{float}
#  - \usepackage[table]{xcolor}
  - \usepackage{setspace}\linespread{1.2}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```

\clearpage
# Model setting

Let $T_{d, s, i}$ denote the time to the $d$-th driver's $s$-th shift's $i$-th critical event. The total number critical events of $d$-th driver's $s$-th shift is $n_{d,s}$. The ranges of these notations are:

- $i = 1, 2, \cdots, n_{d, S_d}$,
- $s = 1, 2, \cdots, S_d$,
- $d = 1, 2, \cdots, D$.

We assume the times of critical events within the $d$-th driver's  $s$-th shift were generated from a non-homogeneous Poisson process (NHPP) with a power law process (PLP), with a fix rate parameter $\beta$ and varying scale parameters $\theta_{d, s}$ across drivers. The data generating process is then:

$$
\begin{aligned}
T_{d, s, 1}, T_{d, s, 2}, \cdots , T_{d, s, n_{d, s}} & \sim \text{PLP}(\beta, \theta_{d, s})\\
\beta & \sim \text{Gamma}(1, 1)\\
\log\theta_{d, s} &= \gamma_{0d} + \gamma_{1}x_{d, s, 1} + \gamma_{2}x_{d, s, 2} + \cdots + \gamma_{k}x_{d, s, k}\\
\gamma_{01}, \gamma_{02}, \cdots, \gamma_{0D} & \sim \text{i.i.d. }N(\mu_0, \sigma_0^2)\\
\gamma_1, \gamma_2, \cdots, \gamma_k & \sim \text{i.i.d. }N(0, 10^2)\\
\mu_0 &\sim N(0, 10^2) \\
\sigma_0 &\sim \text{Gamma}(1, 1)
\end{aligned}
$$

\clearpage
# Simulating data

## Theoretical data generating process (DGP)

1. Random intercepts $\gamma_{01}, \gamma_{02}, \cdots, \gamma_{0D}$. The standard deviation of $\mu_0$ was intentionally set to small number 2 to make $\theta_{d, s}$ fall into a reasonably small range. If I otherwise set it as 10,  $\theta_{d, s}$ may be more than $10^5$ due to the exponentiation, which may not be realistic in real-life data.
$$
\begin{aligned}
\mu_0 = 0 &, \quad \sigma_0 = 0.5 \\
\sigma_0 &\sim \text{Gamma}(1, 1)\\
\gamma_{01}, \gamma_{02}, \cdots, \gamma_{0D} & \sim \text{i.i.d. }N(\mu_0, \sigma_0^2)
\end{aligned}
$$

2. Fixed parameters: 3 fixed parameters $\gamma_1, \gamma_2, \gamma_3$. $$\gamma_1, \gamma_2, \gamma_3 \sim \text{i.i.d. }N(0, 0.5^2)$$

3. The number of observations in the $d$-th driver: $N_{d}$. $$N_{d} \sim \text{Poisson}(10)$$

4. Data: 3 predictor variables $x_{d, s, 1}, x_{d, s, 2}, x_{d, s, 3}$.
$$
\begin{aligned}
x_{d, s, 1} &\sim \text{N}(0, 1)\\
x_{d, s, 2} &\sim \text{Gamma}(1, 1)\\
x_{d, s, 3} &\sim \text{Poisson}(0.2)
\end{aligned}
$$

5. Scale parameters of a NHPP (random effects): $\theta_{d, s}$.
$$
\theta_{d, s} = \text{EXP}(\gamma_{0d} + \gamma_{1}x_{d, s, 1} + \gamma_{2}x_{d, s, 2} + \gamma_{k}x_{d, s, 3})
$$

6. Shape parameter of a NHPP (fixed effect): $\beta  \sim \text{Gamma}(1, 1)$. Set $$\beta = 1.5$$

7. Simulate truncate time $\tau_s$ for each shift. $$\tau_s \sim N(10, 1.3)$$

8. Simulate a NHPP based on $\beta$ and $\theta_{d, s}$. $$T_{d, s, 1}, T_{d, s, 2}, \cdots , T_{d, s, n_{d, s}} \sim \text{PLP}(\beta, \theta_{d, s})$$

\clearpage
\singlespacing
## R code to simulate data and parameters according to the DGP
```{r parameters}
pacman::p_load(rstan, tidyverse, data.table)
source("functions/NHPP_functions.R")

set.seed(123)
D = 10 # the number of drivers
K = 3 # the number of predictor variables

# 1. Random-effect intercepts
# hyperparameters
mu0 = 0
sigma0 = 0.5
r_0D = rnorm(D, mean = mu0, sd = sigma0)

# 2. Fixed-effects parameters
R_K = rnorm(K, mean = 0, sd = 0.5)

# 3. The number of observations (shifts) in the $d$-th driver: $N_{d}$
N_K = rpois(D, 10)
N = sum(N_K) # the total number of obs
id = rep(1:D, N_K)

# 4. Generate data: x_1, x_2, .. x_K
sim1 = function(group_sizes = N_K){
  ntot = sum(group_sizes)
  
  int1 = rep(1, ntot)
  x1 =  rnorm(ntot, 0, 1)
  x2 = rgamma(ntot, 1, 1)
  x3 =  rpois(ntot, 0.2)
  
  return(data.frame(int1, x1, x2, x3))
}
X = sim1(N_K)

# 5. Scale parameters of a NHPP
# 5a. parameter matrix: P
P = cbind(r0 = rep(r_0D, N_K), t(replicate(N, R_K)))
M_logtheta = P*X

# returned parameter for each observed shift
beta = 1.5
theta = exp(rowSums(M_logtheta))
round(theta, 3)
```

## Generate NHPP data to pass to `rstan`

```{r NHPPdata}
sim_hier_plp_tau = function(){
  t_list = list()
  len_list = list()
  tau_vector = rnorm(N, 10, 1.3)
  
  for (i in 1:N) {
    t_list[[i]] = sim_plp_tau(tau_vector[i], beta, theta[i])
    len_list[[i]] = length(t_list[[i]])
  }

  event_dat = data.frame(
    shift_id = rep(1:N, unlist(len_list)),
    event_time = Reduce(c, t_list)
  )

  start_end_dat = data.frame(
    shift_id = 1:N,
    start_time = rep(0, N),
    end_time = tau_vector #difference2
  )
  
  return(list(event_dat = event_dat,
              start_end_dat = start_end_dat,
              shift_length = unlist(len_list)))
}

df = sim_hier_plp_tau()

hier_dat = list(
    N = nrow(df$event_dat),
    K = nrow(df$start_end_dat),
    D = id, #driver index
    tau = df$start_end_dat$end_time,
    event_time = df$event_dat$event_time,
    s = df$shift_length, #the number of events in each shift
    x1 = X[,2], x2 = X[,3], x3 = X[,4]
)
```

\clearpage

# `Stan` code

```{r comment = NA, echo=FALSE}
cat(readLines("stan/nhpp_plp_tau_ML.stan"), sep = c("\n"))
```


# Estimated results
## A single simulation to demonstrate

```{r}
f = stan("stan/nhpp_plp_tau_ML.stan",
         chains = 1, iter = 1000, data = hier_dat, refresh = 0)
```

```{r}
pacman::p_load(magrittr)
est = broom::tidy(f)

pull_est = function(var = "theta", est_obj = f){
  z = est_obj %>% 
    broom::tidy() %>% 
    filter(grepl(var, term)) %>% 
    pull(estimate) %>% 
    round(3)
  return(z)
}
```
\doublespacing

Estimated values: 

- Hyperparameters: $\hat{\mu}_0$: `r pull_est("mu0")`, $\hat{\sigma}_0$: `r pull_est("sigma0")`
- Individual level parameters: $\gamma_1, \gamma_2, \gamma_3$: `r pull_est("r\\[")`
- Rate parameter $\beta$: `r pull_est("beta")`
- $\theta$: `r pull_est("theta")`

## Scale up simulation

To be added.

\clearpage
# Further improvement

In Stan code:

- Need a data matrix $X$,
- Need matrix multiplication,

In data:

- Need a driver index $d = 1, 2, \cdots, K$ for each shift $k$
- Need a data matrix $X$