---
title: "Estimation Results of Bayesian hierarchical NHPP"
author: "Miao Cai <miao.cai@slu.edu>"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2: 
    number_sections: true
    toc: yes
header-includes:
  - \usepackage{soul}
  - \usepackage{float}
#  - \usepackage[table]{xcolor}
  - \usepackage{setspace}\linespread{1.2}
  - \usepackage{lscape}
---

\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 4, fig.pos = "H")
```

\clearpage

```{r}
pacman::p_load(data.table, dplyr, ggplot2, purrr)

read_all <- function(file_path = "fit/hierarchical_NHPP20190713/"){
  est_df = file_path %>% 
    list.files() %>% 
    map_dfr(function(file_name){
      fread(paste0(file_path, file_name)) %>% 
        mutate(id = gsub(".csv", "", file_name)) %>% 
        mutate(id = gsub("_.*", "", id)) %>% 
        mutate(id = gsub("sim", "", id)) %>% 
        mutate(id = as.integer(id))
    }) %>% 
    group_by(term, id) %>% 
    summarise(est_mean = mean(estimate),
              est_sd = sd(estimate)) %>% 
    arrange(term, id)

  return(est_df)
}

plot_est = function(data, var = "beta", hline_var = 1.5){
  p = data %>% 
    filter(term == var) %>% 
    ggplot(aes(id, est_mean)) + 
    geom_point() + 
    geom_line(linetype = "dashed", color = "red")+
    geom_errorbar(aes(ymax = est_mean + 1.96*est_sd,
                      ymin = est_mean - 1.96*est_sd), 
                  width = 1)+
    geom_segment(aes(x = 10, xend = 100, 
                     y = hline_var, yend = hline_var), 
                 color = "green")+
    scale_x_continuous(breaks = c(0, 10, 25, 50, 75, 100),
                       labels = c("0", "10", "25", "50", "75", "100")) +
    labs(x = "The number of drivers (random effects)",
         y = var) +
    theme_bw()
  return(p)
}
```


100 simulation on 2019-07-14
============================

```{r}
est_df1 = read_all("fit/hierarchical_NHPP20190714/")
```

## beta

```{r}
plot_est(est_df1, "beta", 1.5)
```


## mu0

```{r}
plot_est(est_df1, "mu0_true", 0.2)
```


## sigma0

```{r}
plot_est(est_df1, "sigma0", 0.5)
```


## Fixed parameters

```{r}
r1 = plot_est(est_df1, "R1_K[1]", 1)
r2 = plot_est(est_df1, "R1_K[2]", 0.3)
r3 = plot_est(est_df1, "R1_K[3]", 0.2)
```

```{r fig.height=10, fig.width=8}
gridExtra::grid.arrange(r1, r2, r3)
```

3000 simulations
================

```{r}
est_df2 = read_all("fit/hierarchical_NHPP20190715/")
```
## beta

```{r}
plot_est(est_df2, "beta", 1.5)
```

## mu0

```{r}
plot_est(est_df2, "mu0_true", 0.2)
```

## Sigma0

```{r}
plot_est(est_df2, "sigma0", 0.5)
```

## Fixed-effect parameters

```{r}
r1 = plot_est(est_df2, "R1_K[1]", 1)
r2 = plot_est(est_df2, "R1_K[2]", 0.3)
r3 = plot_est(est_df2, "R1_K[3]", 0.2)
```

```{r fig.height=10, fig.width=8}
gridExtra::grid.arrange(r1, r2, r3)
```

\blandscape
Real data estimation
====================

```{r}
rds_path = "fit/50driver_NHPP_fit/"
rds_file = list.files(rds_path)

est_df = rds_path %>% 
  list.files() %>% 
  map_dfr(function(rds_file){
    readRDS(paste0(rds_path, rds_file)) %>% 
      mutate(id = gsub(".rds", "", rds_file)) %>%
      mutate(id = gsub("f_", "", id))
    }) %>% 
  filter(grepl("beta|mu0_true|sigma0|R1", term)) %>% 
  arrange(term, id) %>% 
  mutate(term = case_when(
    term == "R1_K[1]" ~ "ping speed",
    term == "R1_K[2]" ~ "precip intensity",
    term == "R1_K[3]" ~ "precip probability",
    term == "R1_K[4]" ~ "wind speed",
    term == "R1_K[5]" ~ "visibility",
    term == "R1_K[6]" ~ "driver age",
    term == "mu0_true" ~ "$\\mu_0$",
    term == "sigma0" ~ "$\\sigma_0$",
    term == "beta" ~ "$\\beta$",
    TRUE ~ term),
    estimate = round(estimate, 4),
    std.error = round(std.error, 4),
    re = paste0(estimate, " (", std.error, ")")
  ) %>% 
  mutate(term = factor(term, levels = 
    c("$\\beta$", "$\\mu_0$", "$\\sigma_0$", 
      "driver age", "ping speed",  "precip intensity",
      "precip probability", "visibility", "wind speed"))) %>% 
  select(term, id, re) %>% 
  tidyr::spread(key = id, value = re)
```


```{r}
pacman::p_load(kableExtra)
knitr::kable(est_df, format = "latex", booktabs = TRUE, 
             col.names = c("parameters", "drivers 1-50", "drivers 101-150", 
                           "drivers 151-200", "drivers 251-300"),
             escape = F, linesep = "") %>% 
  kable_styling(latex_options = c("scale_down")) %>%
  pack_rows("NHPP parameters", 1, 3) %>%
  pack_rows("Covariate parameters", 4, 9)
```



Potential problems and further improvement

- A eight-hour threshold is not sufficient to separate shifts,
- Should I just delete the extremely long shifts due to the imperfect eight-hour threshold?
- Should I compute the time to event within shifts or trips?
- Imperfect repair between trips and within shifts?
- Bathtube shape intensity?
- Submodels by different types of critical events?

\elandscape