---
title: "Examine perfect fit problem of hierarchical Poisson regression"
author: "Miao Cai <miao.cai@slu.edu>"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2: 
    number_sections: true
    toc: true
header-includes:
  - \usepackage{soul}
  - \usepackage{float}
  - \usepackage[usenames,dvipsnames]{xcolor}
  - \usepackage{setspace}\linespread{1.4}
  - \usepackage{lscape}
  - \usepackage{longtable}
  - \usepackage{booktabs}
---

\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
```

\clearpage
Statistical model
===

$$
\begin{aligned}
Y_i & \sim \text{Poisson}(\lambda_i)\\
\log \lambda_i &= \beta_{0, i} + \beta_1x_1 + \beta_2x_2\\
\beta_{0, i} & \sim N(\mu_0, \sigma_0^2)
\end{aligned}
$$



Simulate data
===

```{r echo = TRUE}
set.seed(123)
N = 50 # total obs
K = N # groups
mu0 = 3 
sigma0 = 1
b0 = rnorm(N, mu0, sigma0)

x1 = rnorm(N, 1, 2)
x2 = rgamma(N, 10, 10)

b1 = 1; b2 = 0.2

lambda = exp(b0 + b1*x1 + b2*x2)

Y = rpois(n = N, lambda = lambda)
```

\clearpage

Stan code
===

```{r comment = NA, echo=FALSE}
cat(readLines("stan/08Perfect_fit_Poisson.stan"), sep = c("\n"))
```

\clearpage

Fit the model
===

```{r}
df = list(
  N = N,
  K = N,
  Y = Y,
  id = 1:N,
  x1 = x1,
  x2 = x2
)

pacman::p_load(rstan)

f = stan("stan/08Perfect_fit_Poisson.stan",
         chains = 1, iter = 1000, data = df, 
         control = list(max_treedepth = 15), refresh = 0)
```

\textcolor{red}{The warning message above highly suggests that there are identifiability issue here in this model since stan cannot sample much effective samples from this model.} The column `n_eff` stands for the number of effective sample size (ESS).  ESS in this model are very very low compared with the number of iterations I used, typically around 30 out of 1000, which means only around 3% of the samples were effective.

This warning message is identical with the following two posts demonstrating the identifiability issues using `Stan`:

- [Identifying non-identifiability by Martin Modrak](https://www.martinmodrak.cz/2018/05/14/identifying-non-identifiability/)
- [Richard McElreath: Statistical Rethinking – 8.4.4. Non-identifiable parameters, page 261](https://bookdown.org/ajkurz/Statistical_Rethinking_recoded/markov-chain-monte-carlo.html#care-and-feeding-of-your-markov-chain.)

```{r}
print(f, probs=c(.1,.5,.9))
```



\clearpage
Compare true and estimated parameters
===

```{r}
f0 = broom::tidy(f)

est = data.frame(
  id = 1:N,
  b0true = b0,
  b0hat = f0$estimate[3:52],
  difference = b0 - f0$estimate[3:52]
)

knitr::kable(est, digits = 3, format = "latex", 
             booktabs = TRUE, escape = F, linesep = "",
             longtable = TRUE,
             col.names = c("id", "$\\beta_0$", 
                           "$\\hat{\\beta}_0$", "$\\Delta$"))
```

There are noticeable difference ($\Delta$) between the true ($\beta_0$) and estimated ($\hat{\beta}_0$) random intercepts.