---
title: "Jump-point PLP (JPLP)"
author: "Miao Cai <miao.cai@slu.edu>"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2: 
    number_sections: true
    toc: true
header-includes:
  - \usepackage{soul}
  - \usepackage{float}
  - \usepackage{graphicx}
  - \usepackage{setspace}\linespread{1.2}
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.pos = 'H', 
                      fig.height = 3.5, fig.align = 'center')
```

# NHPP and PLP

**Intensity function**: The intensity function of a point process is 

$$\lambda(t) = \lim_{\Delta t \rightarrow 0}\frac{P(N(t, t+\Delta t] \geq 1)}{\Delta t}$$

When there is no simultaneous events, ROCOF is the same as intensity function.

**Nonhomogeneous Poisson Process (NHPP)**: The NHPP is a Poisson process whose intensity function is non-constant.

**Power law process (PLP)**: When the intensity function of a NHPP is:

$$\lambda(t) = \frac{\beta}{\theta}\bigg(\frac{t}{\theta}\bigg)^{\beta-1}$$
Where $\beta > 0$ and $\theta > 0$, the process is called the power law process (PLP).

Therefore, the mean function $\Lambda(t)$ is the integral of the intensity function:

$$\Lambda(t) = \int_0^t \lambda(t)dt = \int_0^t \frac{\beta}{\theta}\bigg(\frac{t}{\theta}\bigg)^{\beta-1} = \bigg(\frac{t}{\theta}\bigg)^{\beta}$$

```{r out.width='\\textwidth'}
library(ggplot2)
lambda = function(beta, theta, t) return(beta/theta^beta*t^(beta-1))

plot_plp = function(beta, theta){
  x = seq(0, 10, 0.1)
  lam_y = lambda(beta, theta, x)
  dat = data.frame(lam_y, x)
  p = ggplot(dat, aes(x, lam_y)) + 
    geom_line() + theme_bw() + 
    ggtitle(paste0("beta = ", beta, "; theta = ", theta))
  return(p)
}

plot_plp(1.1, 2)
```

```{r out.width='\\textwidth'}
plot_plp(0.8, 2)
```


\clearpage
# Jump-point PLP (JPLP)

At the time of the failure or rest, the intensity will bounce back at a certain percent $\kappa$, and $0 < \kappa < 1$.

```{r plpnhpp, fig.cap='Intensity function, time to SCEs, and rest time within a shift generated from a NHPP with a PLP intensity function, $\\beta = 1.2$, $\\theta=2$', echo=FALSE, eval=TRUE, out.height='250pt', out.width='\\textwidth'}
knitr::include_graphics("figs/PLP_NHPP_intensity.pdf")
```

```{r plpjumppoint, fig.cap='Intensity function, time to SCEs, and rest time within a shift with a jump-point PLP intensity function, $\\beta = 1.2$, $\\theta=2$, $\\kappa=0.8$', echo=FALSE, eval=TRUE, out.height='250pt', out.width='\\textwidth'}
knitr::include_graphics("figs/PLP_jump_point_intensity.pdf")
```

# Complete intensity function

The complete intensity function of a NHPP with a PLP intensity function is:
$$\lambda^\star(t) = \frac{\beta}{\theta}\bigg(\frac{t}{\theta}\bigg)^{\beta-1}$$

The complete intensity function of a JPLP is then:
\begin{equation}
  \lambda_{JPLP}(t) =
    \begin{cases}
      \lambda^\star(t) & \quad  0 \leq t \leq t_1\\
      \kappa\lambda^\star(t) & \quad t_1 < t \leq t_2\\
      \kappa^2\lambda^\star(t) & \quad t_2 \leq t \leq t_3\\
      \cdots & \cdots\\
      \kappa^{n-1}\lambda^\star(t) & \quad t_{n-1} \leq t \leq t_n\\
      \kappa^{n}\lambda^\star(t) & \quad t_{n} \leq t < \tau
    \end{cases}       
\end{equation}
where $t_1, t_2, \cdots, t_n$ are the time to the 1st, 2nd, $\cdots$, and n-th event. $\tau$ is the truncation time.

Likelihood function
=========================
The first event
---------------
The survival function for the first event is 
$$
\begin{aligned}
S_1(t_1) & = P(T_1 > t_1)\\
         & = \exp\bigg(-\int_0^{t_1}\lambda(x)dx\bigg)\\
         & = \exp\bigg(-\int_0^{t_1}\lambda^\star(x)dx\bigg)
\end{aligned}
$$
The likelihood function is then
$$
\begin{aligned}
f_1(t_1) & = -S_1^\prime(t_1)\\
         & = \lambda^\star(t_1)\exp\bigg(-\int_0^{t_1}\lambda^\star(x)dx\bigg)
\end{aligned}
$$

The second event
----------------
The survival function for the second event is 
$$
\begin{aligned}
S_2(t_2) & = P(T_2 > t_2|t_1)\\
         & = P(N(t_1, t_2]) = 0)\\
         & = \exp\bigg(-\int_{t_1}^{t_2}\kappa\lambda^\star(x)dx\bigg)
\end{aligned}
$$

The associated likelihood function is:
$$
\begin{aligned}
f_2(t_2) & = -S_2^\prime(t_2)\\
         & = \kappa\lambda^\star(t_2)\exp\bigg(-\int_{t_1}^{t_2}\kappa\lambda^\star(x)dx\bigg)
\end{aligned}
$$

The n-th event
--------------
Likewise, the likelihood function of the n-th event is:
$$
\begin{aligned}
f_n(t_n|t_1, t_2, \cdots, t_n) & = -S_n^\prime(t_n)\\
         & = \kappa^{n-1}\lambda^\star(t_n)\exp\bigg(-\int_{t_{n-1}}^{t_n}\kappa^{n-1}\lambda^\star(x)dx\bigg)
\end{aligned}
$$

All events
----------
The likelihood function of all events is:
$$
\begin{aligned}
& f(t_1, t_2, \cdots, t_n) \\
& = f_1(t_1)f_2(t_2|t_1)\cdots f_n(t_n|t_1, t_2,\cdots,t_{n-1})\\
         & = \kappa^0\lambda^\star(t_1)\exp\bigg(-\int_0^{t_1}\lambda^\star(x)dx\bigg)\times \kappa^1\lambda^\star(t_2)\exp\bigg(-\int_{t_1}^{t_2}\kappa\lambda^\star(x)dx\bigg) \times \cdots\times \quad \kappa^{n-1}\lambda^\star(t_n)\exp\bigg(-\int_{t_{n-1}}^{t_n}\kappa^{n-1}\lambda^\star(x)dx\bigg)\\
         & = \kappa^{n(n-1)/2}\bigg(\prod_{i=1}^n\lambda^\star(t_i)\bigg) \times\exp\Bigg(-\bigg[\Big(\frac{t_1}{\theta}\Big)^\beta - \Big(\frac{0}{\theta}\Big)^\beta + \kappa\Big(\frac{t_2}{\theta}\Big)^\beta - \kappa\Big(\frac{t_1}{\theta}\Big)^\beta + \cdots \kappa^{n-1}\Big(\frac{t_n}{\theta}\Big)^\beta - \kappa^{n-1}\Big(\frac{t_{n-1}}{\theta}\Big)^\beta\bigg]\Bigg)\\
         & = \kappa^{n(n-1)/2}\bigg(\prod_{i=1}^n\lambda^\star(t_i)\bigg)\times\exp\Bigg(-\bigg[(1-\kappa)\Big(\frac{t_1}{\theta}\Big)^\beta + \kappa(1-\kappa)\Big(\frac{t_2}{\theta}\Big)^\beta + \cdots + \kappa^{n-2}(1-\kappa)\Big(\frac{t_{n-1}}{\theta}\Big)^\beta + \kappa^{n-1}\Big(\frac{t_{n}}{\theta}\Big)^\beta\bigg]\Bigg)\\
         & = \kappa^{n(n-1)/2}\bigg(\prod_{i=1}^n\lambda^\star(t_i)\bigg)\exp\Bigg(\bigg[\sum_{j=1}^{n-1}\kappa^{j-1}(1-\kappa)\Big(\frac{t_j}{\theta}\Big)^\beta\bigg] + \kappa^{n-1}\Big(\frac{t_n}{\theta}\Big)^\beta\Bigg)
\end{aligned}
$$





















```{r eval=FALSE, echo=FALSE}
plp = function(t, beta = 2, theta = 10) return(beta*theta^(-beta)*t^(beta-1))
sim_plp_tau = function(tau = 30,
                       beta = 2,
                       theta = 10){
  # initialization
   s = 0; t = 0
  while (max(t) <= tau) {
    u <- runif(1)
    s <- s - log(u)
    t_new <- theta*s^(1/beta)
    t <- c(t, t_new)
  }
  t = t[c(-1, -length(t))] 
  
  return(t)
}
```

```{r regularplp, eval=FALSE, echo=FALSE}
pacman::p_load(ggplot2, dplyr, extrafont)

set.seed(223)
tau0 = 13; beta0 = 1.2; theta0 = 2
rest_time = c(4, 5, 8.2, 9.2)

# SCE time
t_SCE = sim_plp_tau(tau = tau0, beta = beta0, theta = theta0)
shift_id= rep(0, length(t_SCE))
SCE = data.frame(shift_id, t_SCE)

# Intensity and limit
INTENS = data.frame(t = seq(0, tau0, 0.001),
                    intensity = plp(seq(0, tau0, 0.001), beta0, theta0)) %>% 
  mutate(type = ifelse(((t >= rest_time[1] & t <= rest_time[2])|(t >= rest_time[3] & t <= rest_time[4])), 0.2, 1))

# LIM data
LIM = data.frame(start_time = 0, end_time = tau0, shift_id = shift_id)

# maximum intensity
max_int = round(max(INTENS$intensity), 2) + 0.01

# Rest
REST1 = data.frame(xaxis = c(rest_time[1:2], rev(rest_time[1:2])),
                   yaxis = rep(c(0, max(INTENS$intensity)), each = 2))
REST2 = data.frame(xaxis = c(rest_time[3:4], rev(rest_time[3:4])),
                   yaxis = rep(c(0, max(INTENS$intensity)), each = 2))

# plot
p0 = ggplot() + 
  geom_line(data = INTENS, aes(x = t, y = intensity), size = 0.8) + 
  geom_point(data = SCE, aes(x = t_SCE, y = shift_id), 
             alpha = 1, shape = 4, color = 'red', size = 2, stroke = 1.5) + 
  geom_segment(data = LIM, 
               aes(x = start_time, xend = end_time, y = shift_id, yend = shift_id),
               arrow = arrow(length = unit(0.2, "cm")), lineend = 'butt', size = 0.8) + 
  scale_x_continuous(expand = c(0.001, 0))+
  scale_y_continuous(labels = c(0, max_int), breaks = c(0, max_int), 
                     expand = c(0.02, 0)) +
  geom_polygon(data = REST1, aes(xaxis, yaxis),
               fill="green", alpha = 0.2) + 
  geom_polygon(data = REST2, aes(xaxis, yaxis),
               fill="green", alpha = 0.2) +
  labs(x = unname(latex2exp::TeX('Time to SCEs (hours), $t_1$ - $t_n$')), 
       y = "intensity") + 
  geom_text(aes(x = c(4.5, 8.7), y = rep(max_int/2, 2), 
                label = c("rest 1","rest 2"))) +
  theme_bw()+
  theme(panel.grid = element_blank())
p0
ggsave("figs/PLP_NHPP_intensity.pdf", p0, width = 10, height = 6.18)
```


```{r plotjumpointPLP, eval=FALSE, echo=FALSE}
set.seed(223)
tau0 = 13; beta0 = 1.2; theta0 = 2; kappa = 0.8

# SCE time
t_SCE = sim_plp_tau(tau = tau0, beta = beta0, theta = theta0)
shift_id= rep(0, length(t_SCE))
SCE = data.frame(shift_id, t_SCE)

# Rest
rest_time = c(4, 5, 8.2, 9.2)
REST1 = data.frame(xaxis = c(rest_time[1:2], rev(rest_time[1:2])),
                   yaxis = rep(c(0,max(INTENS$intensity)), each = 2))
REST2 = data.frame(xaxis = c(rest_time[3:4], rev(rest_time[3:4])),
                   yaxis = rep(c(0,max(INTENS$intensity)), each = 2))

# Intensity and limit
INTENS = data.frame(t = seq(0, tau0, 0.001),
                    intensity = plp(seq(0, tau0, 0.001), beta0, theta0)) %>% 
  mutate(intensity1 = case_when(
    t >= rest_time[2] & t <= rest_time[3] ~ intensity*kappa^1,
    t >= rest_time[4]  ~ intensity*kappa^2,
     TRUE ~ intensity*kappa^0))

# plot
p1 = ggplot() + 
  geom_line(data = INTENS[INTENS$t <= rest_time[1],], 
            aes(x = t, y = intensity1), size = 0.8) + 
  geom_line(data = INTENS[INTENS$t >= rest_time[2]&INTENS$t <= rest_time[3],], 
            aes(x = t, y = intensity1), size = 0.8) + 
  geom_line(data = INTENS[INTENS$t >= rest_time[4],], 
            aes(x = t, y = intensity1), size = 0.8) + 
  geom_line(data = INTENS, 
            aes(x = t, y = intensity), alpha = 0.3, size = 0.8)+
  geom_point(data = SCE, aes(x = t_SCE, y = shift_id), 
             alpha = 1, shape = 4, color = 'red', size = 2, stroke = 1.5) + 
  geom_segment(data = LIM, 
               aes(x = start_time, xend = end_time, y = shift_id, yend = shift_id),
               arrow = arrow(length = unit(0.2, "cm")), lineend = 'butt', size = 0.8) + 
  scale_x_continuous(expand = c(0.001, 0))+
  scale_y_continuous(labels = c(0, max_int), breaks = c(0, max_int), 
                     expand = c(0.02, 0)) +
  geom_polygon(data = REST1, aes(xaxis, yaxis),
               fill="green", alpha = 0.2) + 
  geom_polygon(data = REST2, aes(xaxis, yaxis),
               fill="green", alpha = 0.2) +
  labs(x = unname(latex2exp::TeX('Time to SCEs (hours), $t_1$ - $t_n$')), 
       y = "intensity") + 
  geom_text(aes(x = c(4.5, 8.7), y = rep(max_int/2, 2), 
                label = c("rest 1","rest 2"))) +
  theme_bw()+
  theme(panel.grid = element_blank())#panel.border = element_blank()
p1
ggsave("figs/PLP_jump_point_intensity.pdf", p1, width = 10, height = 6.18)
```


## Updated plot of PLP


```{r REVregularplp, eval=FALSE, echo=FALSE}
pacman::p_load(ggplot2, dplyr, extrafont)
font_import(pattern = "lmroman*")
loadfonts()

plp = function(t, beta = 2, theta = 10) return(beta*theta^(-beta)*t^(beta-1))

set.seed(223)
tau0 = 13; beta0 = 1.2; theta0 = 2
rest_time = c(4, 8)

# SCE time
t_SCE = c(1.5081062, 2.9090437, 3.821632, 
          4.75131, 6.3752535, 7.2168730, 
          10.5380954, 12.1135213)
shift_id= rep(0, length(t_SCE))
SCE = data.frame(shift_id, t_SCE)

# Intensity and limit
INTENS = data.frame(t = seq(0, tau0, 0.001),
                    intensity = plp(seq(0, tau0, 0.001), beta0, theta0)) %>% 
  mutate(type = ifelse(((t >= rest_time[1] & t <= rest_time[2])|(t >= rest_time[3] & t <= rest_time[4])), 0.2, 1))

# LIM data
LIM = data.frame(start_time = 0, end_time = tau0, shift_id = shift_id)

# maximum intensity
max_int = round(max(INTENS$intensity), 2) + 0.01

# plot
p0 = ggplot() + 
  geom_line(data = INTENS, aes(x = t, y = intensity), size = 0.8) + 
  geom_point(data = SCE, aes(x = t_SCE, y = shift_id), 
             alpha = 1, shape = 4, color = 'red', size = 2, stroke = 1.5) + 
  geom_segment(data = LIM, 
               aes(x = start_time, xend = end_time, y = shift_id, yend = shift_id),
               arrow = arrow(length = unit(0.2, "cm")), lineend = 'butt', size = 0.8) + 
  scale_x_continuous(expand = c(0.001, 0))+
  scale_y_continuous(labels = c(0, max_int), breaks = c(0, max_int), 
                     expand = c(0.02, 0)) +
  geom_vline(xintercept = rest_time[1], color="green", 
             alpha = 0.7, size = 1.2, linetype = "dashed") + 
  geom_vline(xintercept = rest_time[2], color="green", 
             alpha = 0.7, size = 1.2, linetype = "dashed") +
  labs(x = unname(latex2exp::TeX('Time to SCEs (hours), $t_1$ - $t_n$')), 
       y = "intensity") + 
  geom_text(aes(x = c(4.5, 8.5), y = rep(max_int/3, 2), 
                label = c("rest 1","rest 2")), 
            color = "darkgreen", size = 5, family="LM Roman 10") +
  theme_bw()+
  theme(panel.grid = element_blank(),
        text = element_text(size=10, family="LM Roman 10"))
p0
ggsave("figs/PLP_NHPP_intensity_rev.pdf", p0, width = 10, height = 6.18)
```


```{r REVJPLP}
set.seed(223)
tau0 = 13; beta0 = 1.2; theta0 = 2; kappa = 0.8
rest_time = c(4, 8)

# SCE time
t_SCE = c(1.5081062, 2.9090437, 3.821632, 
          4.75131, 6.3752535, 7.2168730, 
          10.5380954, 12.1135213)
shift_id= rep(0, length(t_SCE))
SCE = data.frame(shift_id, t_SCE)

# Intensity and limit
INTENS = data.frame(t = seq(0, tau0, 0.001),
                    intensity = plp(seq(0, tau0, 0.001), beta0, theta0)) %>% 
  mutate(intensity1 = case_when(
    t >= rest_time[1] & t <= rest_time[2] ~ intensity*kappa^1,
    t >= rest_time[2]  ~ intensity*kappa^2,
     TRUE ~ intensity*kappa^0))

# plot
p1 = ggplot() + 
  geom_line(data = INTENS[INTENS$t < rest_time[1],], 
            aes(x = t, y = intensity1), size = 0.8) + 
  geom_line(data = INTENS[INTENS$t > rest_time[1]&INTENS$t <= rest_time[2],], 
            aes(x = t, y = intensity1), size = 0.8) + 
  geom_line(data = INTENS[INTENS$t > rest_time[2],], 
            aes(x = t, y = intensity1), size = 0.8) + 
  geom_line(data = INTENS, 
            aes(x = t, y = intensity), alpha = 0.3, size = 0.8)+
  geom_point(data = SCE, aes(x = t_SCE, y = shift_id), 
             alpha = 1, shape = 4, color = 'red', size = 2, stroke = 1.5) + 
  geom_segment(data = LIM, 
               aes(x = start_time, xend = end_time, y = shift_id, yend = shift_id),
               arrow = arrow(length = unit(0.2, "cm")), lineend = 'butt', size = 0.8) + 
  scale_x_continuous(expand = c(0.001, 0))+
  scale_y_continuous(labels = c(0, max_int), breaks = c(0, max_int), 
                     expand = c(0.02, 0)) +
  geom_vline(xintercept = rest_time[1], color="green", 
             alpha = 0.7, size = 1.2, linetype = "dashed") + 
  geom_vline(xintercept = rest_time[2], color="green", 
             alpha = 0.7, size = 1.2, linetype = "dashed")+
  labs(x = unname(latex2exp::TeX('Time to SCEs (hours), $t_1$ - $t_n$')), 
       y = "intensity") + 
  geom_text(aes(x = c(4.5, 8.5), y = rep(max_int/3, 2), 
                label = c("rest 1","rest 2")), 
            color = "darkgreen", size = 5, family="LM Roman 10") +
  theme_bw()+
  theme(panel.grid = element_blank(),
        text = element_text(size=10, family="LM Roman 10"))
p1
ggsave("figs/PLP_jump_point_intensity.pdf", p1, width = 10, height = 6.18)
```

