---
title: "Jump-point PLP (JPLP) simulation"
author: "Miao Cai <miao.cai@slu.edu>"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2: 
    number_sections: true
    toc: true
header-includes:
  - \usepackage{soul}
  - \usepackage{float}
  - \usepackage{graphicx}
  - \usepackage{setspace}\linespread{1.2}
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Power law process (PLP)
=======================
PLP intensity function
----------------------
**Power law process (PLP)**: When the intensity function of a NHPP is:

$$\lambda(t) = \frac{\beta}{\theta}\bigg(\frac{t}{\theta}\bigg)^{\beta-1} = \beta \theta^{-\beta}t^{\beta},$$
where $\beta > 0$ and $\theta > 0$, the process is called the power law process (PLP). The mean function $\Lambda(t)$ is the integral of the intensity function:

$$\Lambda(t) = \int_0^t \lambda(t)dt = \int_0^t \frac{\beta}{\theta}\bigg(\frac{t}{\theta}\bigg)^{\beta-1} = \bigg(\frac{t}{\theta}\bigg)^{\beta}$$

PLP simulation
--------------

```{r}
# simulating PLP - time truncated case
sim_plp_tau = function(tau = 30,
                       beta = 1.5,
                       theta = 10){
  # initialization
  s = 0; t = 0
  while (max(t) <= tau) {
    u <- runif(1)
    s <- s - log(u)
    t_new <- theta*s^(1/beta)
    t <- c(t, t_new)
  }
  t = t[c(-1, -length(t))]

  return(t)
}

# simulate multiple NHPPs - time truncated case
sim_mul_plp_tau = function(n_shift = 20,
                           shift_len_mean = 20, shift_len_sd = 5,
                           theta = 10, beta = 2, mean_n = 5){
  tau_vector = rnorm(n_shift, shift_len_mean, shift_len_sd)#difference1

  t_list = list()
  len_list = list()
  # end_time1 = list() # not needed for time truncated case


  for (i in 1:n_shift) {
    t_list[[i]] = sim_plp_tau(tau_vector[i], beta, theta)
    len_list[[i]] = length(t_list[[i]])
  }

  event_dat = data.frame(
    shift_id = rep(1:n_shift, unlist(len_list)),
    event_time = Reduce(c, t_list)
  )

  start_end_dat = data.frame(
    shift_id = 1:n_shift,
    start_time = rep(0, n_shift),
    end_time = tau_vector #difference2
  )

  return(list(event_dat = event_dat,
              start_end_dat = start_end_dat,
              shift_length = unlist(len_list)))
}
```



```{r}
# plot events
plot_events = function(event_dat, start_end_dat, cross_size = 2){
  p = event_dat %>%
    ggplot(aes(x = event_time, y = shift_id)) +
    geom_point(alpha = 0.8, shape = 4, color = 'red', size = cross_size) +
    scale_y_continuous("shift ID",
                       labels = as.character(start_end_dat$shift_id),
                       breaks = start_end_dat$shift_id)+
    xlab('Time to event (minutes)') +
    geom_segment(data = start_end_dat,
                 aes(x = start_time, xend = end_time,
                     y = shift_id, yend = shift_id),
                 lineend = 'butt',
                 arrow = arrow(length = unit(0.2, "cm"))) +
    theme_classic()
  return(p)
}
```

Jump Power Law Process (JPLP)
=============================
JPLP intensity function
-----------------------
A Bayesian hierarchical JPLP has the following intensity function:
\begin{equation}
\begin{aligned}
  \lambda_{\text{JPLP}}(t|d, s, r, \beta, \gamma_{0, d}, \mathbf{\gamma}, \mathbf{X}_{d}, \mathbf{W}) & =
    \begin{cases}
      \kappa^0\lambda(t|\beta, \gamma_{0, d}, \mathbf{\gamma}, \mathbf{X}_{d}, \mathbf{W}) & \quad  0 \leq t \leq a_{d, s, 1}\\
      \kappa^1\lambda(t|\beta, \gamma_{0, d}, \mathbf{\gamma}, \mathbf{X}_{d}, \mathbf{W}) & \quad  a_{d, s, 1} \leq t \leq a_{d, s, 2}\\
      \cdots & \cdots\\
      \kappa^{R-1}\lambda(t|\beta, \gamma_{0, d}, \mathbf{\gamma}, \mathbf{X}_{d}, \mathbf{W}) & \quad  a_{d, s, R-1} \leq t \leq a_{d, s, R}
    \end{cases} \\
    & = \kappa^{r-1}\lambda(t|d, s, r, \kappa, \beta, \gamma_{0, d}, \mathbf{\gamma}, \mathbf{X}_{d}, \mathbf{W}) \quad  a_{d, s, r-1} \leq t \leq a_{d, s, r},
\end{aligned}
\end{equation}
where the introduced parameter $\kappa$ is the percent of intensity function recovery once the driver takes a break. We assume that this $\kappa$ is constant across drivers and shifts.


JPLP mean function
------------------
> Let $T_1, T_2, \ldots$ be random variables representing the event times of a nonhomogeneous Poisson process with continuous expectation function $\Lambda(t)$, and let $N_t$ represent the total number of events occurring before time $t$ in the process. Then, conditional on the number of events $N_{t_0} = n$, the event times $T_1, T_2, \ldots, T_n$ are distributed as order statistics from a sample with distribution function $F(t) = \Lambda(t)/\Lambda(t_0)$ for $t \in [0, t_0]$.

This is a generalization of the result for homogeneous Poisson processes. It naturally gives rise to the following algorithm for generating random variates from a nonhomogeneous Poisson process with expectation function $\Lambda(t)$ in a fixed interval $[0, t_0]$.

(1) Generate $n \sim \text{Poisson}(\Lambda(t_0))$.
(2) Independently generate $n$ random variates $t_1^\prime, t_2^\prime, \ldots, t_n^\prime$ from the cdf $F(t) = \Lambda(t)/\Lambda(t_0)$.
(3) Order $t_1^\prime, t_2^\prime, \ldots, t_n^\prime$ to obtain $t_1 = t_{(1)}^\prime, t_2 = t_{(2)}^\prime, \ldots, t_n = t_{(n)}^\prime$.
(4) Deliver $t_1, t_2, \ldots, t_n$.


### Mean function $\Lambda(t)$
```{r}
# Mean function Lambda for JPLP
L_plp = function(t, beta = 1.5, theta = 4) return((t/theta)^beta)
Lambda = function(t,
                  tau = 12,
                  kappa = 0.8,
                  t_trip = c(3.5, 6.2, 9),
                  beta = 1.5,
                  theta = 4)
{
  t_trip1 = c(0, t_trip)
  n_trip = length(t_trip1)
  comp = L_plp(t_trip, beta, theta)
  kappa_vec0 = rep(kappa, n_trip - 1)^(0:(n_trip - 2))
  kappa_vec1 = rep(kappa, n_trip - 1)^(1:(n_trip - 1))
  cum_comp0 = comp*kappa_vec0
  cum_comp1 = comp*kappa_vec1
  index_trip = max(cumsum(t > t_trip1)) - 1
  
  if(index_trip == 0){
    return((t/theta)^beta)
  }else{
    return(sum(cum_comp0[1:index_trip]) - sum(cum_comp1[1:index_trip]) +
             kappa^index_trip*(t/theta)^beta)
  }
}
```

### Test the mean function $\Lambda(t)$

```{r}
# test Lambda
kappa = 0.8
t_trip = c(3.5, 6.2, 9)
beta = 1.5
theta = 4

Lambda(3.1)
L_plp(3.1)

Lambda(4.1)
kappa^0*L_plp(t_trip[1]) + 
  kappa^1*L_plp(4.1) - kappa^1*L_plp(t_trip[1])

Lambda(8.9)
kappa^0*L_plp(t_trip[1]) + 
  kappa^1*L_plp(t_trip[2]) - kappa^1*L_plp(t_trip[1]) +
  kappa^2*L_plp(8.9) - kappa^2*L_plp(t_trip[2])

Lambda(12)
kappa^0*L_plp(t_trip[1]) + 
  kappa^1*L_plp(t_trip[2]) - kappa^1*L_plp(t_trip[1]) +
  kappa^2*L_plp(t_trip[3]) - kappa^2*L_plp(t_trip[2]) + 
  kappa^3*L_plp(12) - kappa^3*L_plp(t_trip[3])
```

### Plot the mean function of PLP and JPLP

```{r}
pacman::p_load(ggplot2, dplyr, tidyr, directlabels)

t_trip = c(3.5, 6.2, 9)
beta = 1.5
theta = 4

x = seq(0.01, 12, 0.01)
y0 = (x/theta)^beta
y1 = rep(NA_real_, length(x))

for (i in 1:length(x)) {
  y1[i] = Lambda(t = x[i], kappa = 0.8)
}

data.frame(x, y0, y1) %>% 
  tidyr::pivot_longer(cols = c('y0', 'y1'), 
                      names_to = "Type", 
                      values_to = 'y') %>% 
  mutate(Type = case_when(Type == 'y0' ~ 'PLP',
                          Type == 'y1' ~ 'JPLP')) %>% 
  mutate(Type = factor(Type, levels = c('PLP', 'JPLP'))) %>% 
  ggplot(aes(x = x, y = y, group = Type, color = Type)) + 
  geom_line() + 
  geom_dl(aes(label = Type),
          method = list(dl.trans(x = x - 1.3, y = y - 0.95), 
                        "last.points", cex = 1.2)) + 
  scale_color_manual(values = c(PLP = "#377eb8", JPLP = "#e41a1c")) + 
  geom_vline(xintercept = t_trip, linetype = "dashed", color = '#4daf4a') +
  guides(color = FALSE) +
  labs(x = 'Time (hours)', 
       y = 'Mean function (Lambda)') + 
  theme_minimal() + 
  scale_x_continuous(expand = c(0, 0), breaks = seq(0, 12, 3))
```


Simulation JPLP events in one shift
-----------------------------------

### Simulate JPLP events in one shift

```{r}
# Define a inverse function for mean function Lambda
inverse = function (f, lower = 0.0001, upper = 100) {
   function (y) uniroot((function (x) f(x) - y), lower = lower, upper = upper)[1]
}

# Test the inverse function
Lambda1 = function(t, tau = 12, kappa = 0.8, t_trip = c(3.5, 6.2, 9),
                beta = 1.5, theta = 4){
  return(Lambda(t, tau = tau, kappa = kappa, t_trip = t_trip,
                beta = beta, theta = theta))
}
inv_Lambda = inverse(Lambda1, 0.0001, 100)
inv_Lambda(3.564885)$root
```

```{r}
# simulating JPLP - time truncated case
sim_jplp = function(tau0 = 12,
                    kappa0 = 0.8,
                    t_trip0 = c(3.5, 6.2, 9),
                    beta0 = 1.5,
                    theta0 = 0.5)
{ s = 0; t = 0
  Lambda1 = function(t, tau1 = tau0, kappa1 = kappa0, t_trip1 = t_trip0,
                     beta1 = beta0, theta1 = theta0){
    return(Lambda(t, tau = tau1, kappa = kappa1, t_trip = t_trip1,
                  beta = beta1, theta = theta1))
  }
  inv_Lambda = inverse(Lambda1, 0.0001, 100)

  while (max(t) <= tau0) {
    u <- runif(1)
    s <- s - log(u)
    t_new <- inv_Lambda(s)$root
    t <- c(t, t_new)
  }
  t = t[c(-1, -length(t))]

  return(t)
}

sim_jplp(theta0 = 1)
sim_jplp(theta0 = 2)
```


### Plot JPLP events in one shift

```{r}
set.seed(123)

tauX = 12
kappaX = 0.8
t_tripX = c(3.5, 6.2, 9)
betaX = 1.5
thetaX = 2

t_events = sim_jplp(tau0 = tauX,
                    kappa0 = kappaX,
                    t_trip0 = t_tripX,
                    beta0 = betaX,
                    theta0 = thetaX)

x = seq(0.01, 12, 0.01)
y0 = (x/theta)^beta
y1 = rep(NA_real_, length(x))

for (i in 1:length(x)) {
  y1[i] = Lambda(t = x[i], kappa = 0.8)
}

dLambda = data.frame(x, y0, y1) %>% 
  tidyr::pivot_longer(cols = c('y0', 'y1'), 
                      names_to = "Type", 
                      values_to = 'y') %>% 
  mutate(Type = case_when(Type == 'y0' ~ 'PLP',
                          Type == 'y1' ~ 'JPLP')) %>% 
  mutate(Type = factor(Type, levels = c('PLP', 'JPLP')))

d_event = data.frame(t_events = t_events, y = 0)
d_shift = data.frame(start_x = 0,
                     end_x = tauX,
                     start_y = 0,
                     end_y = 0)
```

```{r fig.cap='Event times (red cross) and mean function Lambda (curves)'}
ggplot() + 
  geom_line(data = dLambda, aes(x = x, y = y, group = Type, color = Type)) + 
  geom_dl(data = dLambda, aes(x = x, y = y, color = Type, label = Type),
          method = list(dl.trans(x = x - 1.3, y = y - 0.9), 
                        "last.points", cex = 1.2)) + 
  scale_color_manual(values = c(PLP = "#377eb8", JPLP = "#e41a1c")) + 
  guides(color = FALSE) +
  geom_vline(xintercept = t_trip, linetype = "dashed", color = '#4daf4a') +
  labs(x = 'Time (hours)', 
       y = 'Mean function (Lambda)') + 
  theme_minimal() + 
  geom_point(data = d_event, aes(x = t_events, y = y), 
             alpha = 0.8, shape = 4, color = 'red', size = 2) + 
  geom_segment(data = d_shift, 
                 aes(x = start_x, xend = end_x, 
                     y = start_y, yend = end_y),
                 lineend = 'butt',
                 arrow = arrow(length = unit(0.2, "cm"))) + 
  scale_x_continuous(expand = c(0, 0), breaks = seq(0, 12, 3))
```

```{r}
set.seed(666)

tauX = 12
kappaX = 0.8
t_tripX = c(3.5, 6.2, 9)
betaX = 1.5
thetaX = 1

t_events = sim_jplp(tau0 = tauX,
                    kappa0 = kappaX,
                    t_trip0 = t_tripX,
                    beta0 = betaX,
                    theta0 = thetaX)

lambda_plp = function(t, beta, theta) return(beta*theta^(-beta)*t^(beta - 1))
lambda_jplp = function(t, beta, theta, kappa = 0.8, t_trip){
  index_trip = max(cumsum(t > t_trip))
  return((kappa^index_trip)*beta*theta^(-beta)*t^(beta - 1))
}


x = seq(0.01, 12, 0.01)
y_plp = lambda_plp(x, beta = betaX, theta = thetaX)
y_jplp = rep(NA_real_, length(lambda_plp))

for (i in 1:length(x)) {
  y_jplp[i] = lambda_jplp(x[i], 
                          beta = betaX, theta = thetaX, 
                          kappa = 0.8, t_trip = t_tripX)
}


dlambda = tibble::tibble(x, y_plp, y_jplp) %>% 
  tidyr::pivot_longer(cols = c('y_plp', 'y_jplp'), 
                      names_to = "Type", 
                      values_to = 'y') %>% 
  mutate(Type = case_when(Type == 'y_plp' ~ 'PLP',
                          Type == 'y_jplp' ~ 'JPLP')) %>% 
  mutate(Type = factor(Type, levels = c('PLP', 'JPLP')))

d_event = data.frame(t_events = t_events, y = 0)
d_shift = data.frame(start_x = 0,
                     end_x = tauX,
                     start_y = 0,
                     end_y = 0)
```

```{r fig.cap='Event times (red cross) and intensity function lambda (curves)'}
ggplot() + 
  geom_line(data = dlambda, aes(x = x, y = y, group = Type, color = Type)) + 
  scale_color_manual(values = c(PLP = "#0B775E", JPLP = "#F2300F")) + 
  geom_dl(data = dlambda, aes(x = x, y = y, label = Type, color = Type),
          method = list(dl.trans(x = x - 1.3, y = y - 0.5), 
                        "last.points", cex = 1.2)) +
  geom_vline(xintercept = t_trip, linetype = "dashed") +
  labs(x = 'Time (hours)', 
       y = 'Mean function (Lambda)') + 
  theme_minimal() + 
  guides(color = FALSE) +
  geom_point(data = d_event, aes(x = t_events, y = y), 
             alpha = 0.8, shape = 4, color = 'red', size = 2) + 
  geom_segment(data = d_shift, 
                 aes(x = start_x, xend = end_x, 
                     y = start_y, yend = end_y),
                 lineend = 'butt',
                 arrow = arrow(length = unit(0.2, "cm"))) + 
  scale_x_continuous(expand = c(0, 0), breaks = seq(0, 12, 3))
```

Simulation JPLP events in multiple shifts
-----------------------------------------

```{r}
# Function to get the number of stops
get_n_stop = function() sample(1:4, 1, TRUE)

# Function to simulate event time for multiple shifts
sim_mul_jplp = function(kappa = 0.8, beta = 1.5, theta = 2, n_shift = 10)
{
  t_shift_vec = list()
  n_stop_vec = list()
  t_stop_vec = list()
  n_event_vec = list()
  t_event_vec = list()
  
  for (i in 1:n_shift) {
    sim_tau = round(runif(1, min = 6, max = 12), 1)
    n_stop = get_n_stop()
    sim_t_trip = round((1:n_stop)*sim_tau/(n_stop + 1) + 
                         rnorm(n_stop, 0, sim_tau*0.15/n_stop), 2)
    t_events = sim_jplp(tau0 = sim_tau,
                        kappa0 = kappa,
                        t_trip0 = sim_t_trip,
                        beta0 = beta,
                        theta0 = theta)
    t_shift_vec[[i]] = sim_tau
    n_stop_vec[[i]] = n_stop
    t_stop_vec[[i]] = sim_t_trip
    n_event_vec[[i]] = length(t_events)
    t_event_vec[[i]] = t_events
  }
  
  event_dt = data.frame(
    shift_id = rep(1:n_shift, unlist(n_event_vec)),
    event_time = Reduce(c, t_event_vec)
  )
  
  trip_dt = data.frame(
    shift_id = rep(1:n_shift, unlist(n_stop_vec)),
    trip_time = Reduce(c, t_stop_vec)
  )

  shift_dt = data.frame(
    shift_id = 1:n_shift,
    start_time = rep(0, n_shift),
    end_time = Reduce(c, t_shift_vec)
  )

  return(list(event_time = event_dt,
              trip_time = trip_dt,
              shift_time = shift_dt))
  
}

set.seed(666)
z = sim_mul_jplp(theta = 3)
str(z)
```

```{r fig.cap='Simulated events (red cross) in different trips (green cross) and shifts'}
# Plot events in multiple shifts
plot_jplp = function(dt){
  p = ggplot() + 
    geom_point(data = dt$event_time, aes(x = event_time, y = shift_id), 
               alpha = 0.8, shape = 4, color = '#F2300F', size = 3, stroke = 1) + 
    geom_point(data = dt$trip_time, aes(x = trip_time, y = shift_id), 
               shape = 3, color = '#0B775E', fill = '#0B775E', size = 3.5, stroke = 1.05) + 
    geom_segment(data = dt$shift_time,
                 aes(x = start_time, xend = end_time,
                     y = shift_id, yend = shift_id),
                 lineend = 'butt',
                 arrow = arrow(length = unit(0.2, "cm"))) +
    scale_y_continuous("shift ID",
                       labels = as.character(dt$shift_time$shift_id),
                       breaks = dt$shift_time$shift_id) + 
    labs(x = 'Time (hours)')
    theme_classic()
    return(p)
}

plot_jplp(z)
```

