---
title: "Hierarchical Jump-point PLP (JPLP) simulation"
author: "Miao Cai <miao.cai@slu.edu>"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2: 
    number_sections: true
    toc: true
header-includes:
  - \usepackage{soul}
  - \usepackage{float}
  - \usepackage{graphicx}
  - \usepackage{setspace}\linespread{1.2}
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Bayesian Hierarchical Jump Power Law Process (JPLP)
===================================================
Model setting
-------------
The Bayesian hierarchical JPLP model is parameterized as
\begin{equation}\label{eq:jplp}
\begin{aligned}
  t_{d, s, 1}, t_{d, s, 2}, \cdots , t_{d, s, n_{d, s}} & \sim \text{JPLP}(\beta, \theta_{d, s}, \kappa)\\
  \beta & \sim \text{Gamma}(1, 1)\\
  \log\theta_{d, s} &= \gamma_{0d} + \gamma_{1}x_{d, s, 1} + \gamma_{2}x_{d, s, 2} + \cdots + \gamma_{k}x_{d, s, k}\\
  \kappa & \sim \text{Uniform}(0, 1)\\
  \gamma_{01}, \gamma_{02}, \cdots, \gamma_{0D} & \sim \text{i.i.d. }N(\mu_0, \sigma_0^2)\\
  \gamma_1, \gamma_2, \cdots, \gamma_k & \sim \text{i.i.d. }N(0, 10^2)\\
  \mu_0 &\sim N(0, 5^2) \\
  \sigma_0 &\sim \text{Gamma}(1, 1),
\end{aligned}
\end{equation}
where the introduced parameter $\kappa$ is the percent of intensity function recovery once the driver takes a break. By definition, $a_{d, s, 0} = 0$. We assume that this $\kappa$ is constant across drivers and shifts.



Intensity function of JPLP
--------------------------
Since the Bayesian hierarchical PLP does not account for the rests within a shift and associated potential reliability repairment. In this subsection, we proposes a Bayesian hierarchical JPLP, with the following intensity function:
\begin{equation}
\begin{aligned}
  \lambda_{\text{JPLP}}(t|d, s, r, \beta, \gamma_{0, d}, \mathbf{\gamma}, \mathbf{X}_{d}, \mathbf{W}) & =
    \begin{cases}
      \kappa^0\lambda(t|\beta, \gamma_{0, d}, \mathbf{\gamma}, \mathbf{X}_{d}, \mathbf{W}) & \quad  0 < t \leq a_{d, s, 1}\\
      \kappa^1\lambda(t|\beta, \gamma_{0, d}, \mathbf{\gamma}, \mathbf{X}_{d}, \mathbf{W}) & \quad  a_{d, s, 1} < t \leq a_{d, s, 2}\\
      \cdots & \cdots\\
      \kappa^{R-1}\lambda(t|\beta, \gamma_{0, d}, \mathbf{\gamma}, \mathbf{X}_{d}, \mathbf{W}) & \quad  a_{d, s, R-1} < t \leq a_{d, s, R}
    \end{cases} \\
    & = \kappa^{r-1}\lambda(t|d, s, r, \kappa, \beta, \gamma_{0, d}, \mathbf{\gamma}, \mathbf{X}_{d}, \mathbf{W}) \quad  a_{d, s, r-1} < t \leq a_{d, s, r},
\end{aligned}
\end{equation}
The notations are identical with those in PLP except for the extra $\kappa$ parameter.

The likelihood function of JPLP
-------------------------------
The likelihood function for driver $d$ on shift $s$ is
\begin{equation}
\begin{split}
    L_{s, d}(\kappa, \beta, \gamma_{0, d}, \mathbf{\gamma}|\text{Data}_{d, s}) = \Big( 
    \prod_{i=1}^{c_{d, s}}\lambda\big( t_{i, d, s}| d, s, r, k, \beta, \gamma_{0, d}, \mathbf{\gamma}, \mathbf{X}_{d}, \mathbf{W} \big)
    \Big)\\ \times \exp\Big(-\int_0^{a_{d, s, r}\lambda\big(u|d, s, r, k, \beta, \gamma_{0, d}, \mathbf{\gamma}, \mathbf{X}_{d}, \mathbf{W}\big)du}\Big)
\end{split}
\end{equation}

The overall likelihood function is
\begin{equation}
    L = \displaystyle\prod_d\prod_{s \in d}L_{s, d}
\end{equation}



Simulating parameters and data
==============================

* Parameters needed: $\kappa, \beta, \theta, \gamma_{0, d}, \mathbf{\gamma}$
    - $\theta \leftarrow \gamma_{0, d}, \mathbf{\gamma}, \mathbf{X}$
* Data needed: $\mathbf{X}$
    - $x_1, x_2, x_3$



```{r}
pacman::p_load(rstan, tidyverse, data.table)
source("functions/JPLP_functions.R")

sim_hier_JPLP = function(
  beta = 1.2,
  kappa = 0.8,
  D = 10, # the number of drivers
  K = 3, # the number of predictor variables
  group_size_lambda = 10, # the mean number of shifts for each driver
  mu0 = 0.2, # hyperparameter 1
  sigma0 = 0.5, # hyperparameter 2
  R_K = c(1, 0.3, 0.2) # Fixed-effects parameters
)
{
  # 1. Random-effect intercepts
  r_0D = rnorm(D, mean = mu0, sd = sigma0)
  
  # 3. The number of observations (shifts) in the $d$-th driver: $N_{d}$
  N_K = rpois(D, group_size_lambda)
  N = sum(N_K) # the total number of shifts for all D drivers
  id = rep(1:D, N_K)

  # 4. Generate data: x_1, x_2, .. x_K
  simX = function(group_sizes = N_K)
    {
      ntot = sum(group_sizes)
      
      int1 = rep(1, ntot)
      x1 =  rnorm(ntot, 1, 1)
      x2 = rgamma(ntot, 1, 1)
      x3 =  rpois(ntot, 2)
      
      return(data.frame(int1, x1, x2, x3))
    }
  X = simX(N_K)
  
  # 5. Scale parameters of a JPLP
  # 5a. parameter matrix: P
  P = cbind(r0 = rep(r_0D, N_K), t(replicate(N, R_K)))
  M_logtheta = P*X
  theta = exp(rowSums(M_logtheta))

  # Initialization of lists
  t_shift_vec = list()
  n_stop_vec = list()
  t_stop_vec = list()
  n_event_vec = list()
  t_event_vec = list()

  for (i in 1:N) 
    {
      sim_tau = rnorm(1, 10, 1.3)
      n_stop = get_n_stop()
      sim_t_trip = round((1:n_stop)*sim_tau/(n_stop + 1) + 
                           rnorm(n_stop, 0, sim_tau*0.15/n_stop), 2)
      t_events = sim_jplp(tau0 = sim_tau,
                          kappa0 = kappa,
                          t_trip0 = sim_t_trip,
                          beta0 = beta,
                          theta0 = theta[i])
      t_shift_vec[[i]] = sim_tau
      n_stop_vec[[i]] = n_stop
      t_stop_vec[[i]] = sim_t_trip
      n_event_vec[[i]] = length(t_events)
      t_event_vec[[i]] = t_events
    }

  

  # shifts data
  shift_dt = data.frame(
    driver_id = rep(1:D, N_K),
    shift_id = 1:N,
    start_time = rep(0, N),
    end_time = Reduce(c, t_shift_vec),
    n_stop = Reduce(c, n_stop_vec),
    n_event = Reduce(c, n_event_vec)
  )
  
  # trips data set
  trip_dt = data.frame(
    driver_id = rep(shift_dt$driver_id, shift_dt$n_stop),
    shift_id = rep(1:N, unlist(n_stop_vec)),
    trip_time = Reduce(c, t_stop_vec)
  )
  
  # TEMPORARY vector: a temporary vector for events per driver
  n_event_driver = shift_dt %>% 
    group_by(driver_id) %>% 
    summarise(n_event = sum(n_event)) %>% 
    pull(n_event)
  
  # events data set
  event_dt = data.frame(
    driver_id = rep(1:D, n_event_driver),
    shift_id = rep(1:N, Reduce(c, n_event_vec)),
    event_time = Reduce(c, t_event_vec)
  )

  return(list(event_time = event_dt,
              trip_time = trip_dt,
              shift_time = shift_dt))
}

set.seed(123)
df = sim_hier_JPLP()
```

Simulated parameters:

- $\kappa$: 
- $\beta$: 
- $\theta$: 
- $\gamma_{0, d}$: 
- $\mathbf{\gamma}$: 

Simulated data:

```{r}
str(df)
```



Generate data pass on to Stan
=============================

```{r JPLPStandata}
# This version is a template
# Need trip time data
sim_hier_nhpp = function(group_size_lambda = 10, D = 10, K = 3, beta = 1.5)
{
  # 1. Random-effect intercepts
  # hyperparameters
  mu0 = 0.2
  sigma0 = 0.5
  r_0D = rnorm(D, mean = mu0, sd = sigma0)

  # 2. Fixed-effects parameters
  R_K = c(1, 0.3, 0.2)

  # 3. The number of shifts in the $d$-th driver: $N_{d}$
  N_K = rpois(D, group_size_lambda)
  N = sum(N_K) # the total number of obs
  id = rep(1:D, N_K)

  # 4. Generate data: x_1, x_2, .. x_K
  sim1 = function(group_sizes = N_K)
  {
    ntot = sum(group_sizes)

    int1 = rep(1, ntot)
    x1 = rnorm(ntot, 1, 1)
    x2 = rgamma(ntot, 1, 1)
    x3 = rpois(ntot, 2)

    return(data.frame(int1, x1, x2, x3))
  }
  X = sim1(N_K)

  # 5. Scale parameters of a NHPP
  # 5a. parameter matrix: P
  P = cbind(r0 = rep(r_0D, N_K),
            t(replicate(N, R_K)))
  M_logtheta = P*X

  # returned parameter for each observed shift

  theta_vec = exp(rowSums(M_logtheta))

  df = sim_hier_plp_tau(N = N, beta = beta, theta = theta_vec)

  hier_dat = list(
    N = nrow(df$event_dat),
    K = K,
    S = nrow(df$start_end_dat),
    D = max(id),
    id = id, #driver index
    tau = df$start_end_dat$end_time,
    event_time = df$event_dat$event_time,
    group_size = df$shift_length, #the number of events in each shift
    X_predictors = X[,2:4]
  )

  true_params = list(
    mu0 = mu0, sigma0 = sigma0,
    r0 = r_0D, r1_rk = R_K,
    beta = beta,
    theta = theta_vec
  )

  return(list(hier_dat = hier_dat, true_params = true_params))
}
```


