---
title: "Bayesian hierarchical models for NHPP using `rstan`"
author: "Miao Cai <miao.cai@slu.edu>"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2: 
    number_sections: true
    toc: true
header-includes:
  - \usepackage{soul}
  - \usepackage{float}
#  - \usepackage[table]{xcolor}
  - \usepackage{setspace}\doublespacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Model setting

Let $T_{d, s, i}$ denote the time to the $d$-th driver's $s$-th shift's $i$-th critical event. The total number critical events of $d$-th driver's $s$-th shift is $n_{d,s}$. The ranges of these notations are:

- $i = 1, 2, \cdots, n_{d, S_d}$,
- $s = 1, 2, \cdots, S_d$,
- $d = 1, 2, \cdots, D$.

We assume the times of critical events within the $d$-th driver's  $s$-th shift were generated from a non-homogeneous Poisson process (NHPP) with a power law process (PLP), with a fix rate parameter $\beta$ and varying scale parameters $\theta_{d, s}$ across drivers. The data generating process is then:

$$
\begin{aligned}
T_{d, s, 1}, T_{d, s, 2}, \cdots , T_{d, s, n_{d, s}} & \sim \text{PLP}(\beta, \theta_{d, s})\\
\beta & \sim \text{Gamma}(1, 1)\\
\log\theta_{d, s} &= \gamma_{0d} + \gamma_{1}x_{d, s, 1} + \gamma_{2}x_{d, s, 2} + \cdots + \gamma_{k}x_{d, s, k}\\
\gamma_{01}, \gamma_{02}, \cdots, \gamma_{0D} & \sim \text{i.i.d. }N(\mu_0, \sigma_0^2)\\
\gamma_1, \gamma_2, \cdots, \gamma_k & \sim \text{i.i.d. }N(0, 10^2)\\
\mu_0 &\sim N(0, 10^2) \\
\sigma_0 &\sim \text{Gamma}(1, 1)
\end{aligned}
$$

# Simulating data

1. Random intercepts $\gamma_{0i}, \gamma_{0i}, \cdots, \gamma_{0D}$.
$$
\begin{aligned}
\mu_0 &\sim N(0, 10^2) \\
\sigma_0 &\sim \text{Gamma}(1, 1)\\
\gamma_{01}, \gamma_{02}, \cdots, \gamma_{0D} & \sim \text{i.i.d. }N(\mu_0, \sigma_0^2)
\end{aligned}
$$

2. fixed parameters: 3 fixed parameters $\gamma_1, \gamma_2, \gamma_3$.
$$
\gamma_1, \gamma_2, \gamma_3 \sim \text{i.i.d. }N(0, 10^2)
$$

3. Data: three variables $x_{d, s, 1}, x_{d, s, 2}, x_{d, s, 3}$.
$$
\begin{aligned}
x_{d, s, 1} &\sim \text{N}(0, 10)\\
x_{d, s, 2} &\sim \text{Gamma}(10, 2)\\
x_{d, s, 3} &\sim \text{Poisson}(3.5)
\end{aligned}
$$

4. scale parameterS of a NHPP (random effects): $\theta_{d, s}$.
$$
\theta_{d, s} = \text{EXP}(\gamma_{0d} + \gamma_{1}x_{d, s, 1} + \gamma_{2}x_{d, s, 2} + \gamma_{k}x_{d, s, 3})
$$

5. Shape parameter of a NHPP (fixed effect): $\beta$.
$$
\beta \sim \text{Gamma}(1, 1)
$$

6. Simulate a NHPP based on $\beta$ and $\theta_{d, s}$.
$$
T_{d, s, 1}, T_{d, s, 2}, \cdots , T_{d, s, n_{d, s}} \sim \text{PLP}(\beta, \theta_{d, s})
$$

```{r eval=FALSE}
set.seed(123)
D = 10 # the number of drivers
K = 3 # the number of predictor variables

# 1. Random intercepts
# hyperparameters
mu0 = rnorm(1, mean = 0, sd = 10)
sigma0 = rgamma(1, 1, 1)


# Fixed-effects parameters
R = rnorm(K, mean = 0, sd = 10)

# Random intercepts
r0 = rnorm(D, mean = mu0, sd = sigma0)

# Generate x_1, x_2, .. x_K
sim1d = function(n = 10){
  
}
x1 = rnorm()
x2 = rgamma()
x3 = rpois()
```


```{r eval=FALSE}
y = c(0, 1, 0, 0, 0, 1, 0, 1, 0, 0, 1, 0, 1)
x = c(1, 3, 2, 0, -1, 5, 2, 6, 2, 5, 3, 3, 1)
fit = glm(y ~ x, family = "binomial")
yhat = predict(fit, type = "response")

sum(y)
sum(yhat)
```

